### 2025-01-20 paper/icml25/first/sects/experiments.tex
first draft for results

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios for the Transformer Architecture.}
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Norm of policy gradients on the Transport environment across 5000 epochs. The raise in norm for SHAC (epoch 2700) corresponds with occurence of generalization, where collision between agents and the packages become frequent. Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurence of generalization (epoch 3200).}
\end{figure}

@@

\begin{table*}[!t]
    \centering
    \input tables/max.tex
    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios.}
\end{table*}

@@

\subsection{Results}

\paragraph{Dispersion}
Firstly, we compare the performance of these algorithms on a non-differentiable scenario (thus SHAC is not aplicable), dispersion. We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture (see \Cref{apx:fig:experiments-mlp}). In single agent setting, show in \Cref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes. With $3$ agents, shown in \Cref{fig:dispersion-transformer-3}, PPO fails to converge, meanwhile \fname{} converges in less than $10,000$ episodes. With $5$ agents, shown in \Cref{fig:dispersion-transformer-5}, PPO fails to converge, meanwhile \fname{} achieves high rewards without however triggering the early stopping. In this last scenario, \fname{} manifest also high reward variance.

\paragraph{Transport}
Now, we compare the performance of these algorithms on a fully differentiable scenario, transport. On the single agent setting, shown in \Cref{fig:transport-mlp-1}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well. While \fname{} appears to converge earlier SHAC appear to be more consistent. With $3$ and $5$ agents, shown in \Cref{fig:transport-transformer-3} and \Cref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable. For the MLP only setting see \Cref{apx:fig:experiments-mlp}

\paragraph{Discovery}
Similar consideration hold also for the discovery scenario. In the single agent setting, shown in \Cref{fig:discovery-mlp-1}, \fname{} performs better, but comparitavely, to PPO . However, with $3$ and $5$ agents, shown in \Cref{fig:discovery-transformer-3} and \Cref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better. For the MLP only setting see \Cref{apx:fig:experiments-mlp}

\paragraph{Sampling}

>>>>

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of normalized rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Norm of policy gradients on the Transport environment across 5000 epochs. The raise in norm for SHAC (epoch 2700) corresponds with occurence of generalization, where collision between agents and the packages become frequent. Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
\end{figure}

@@

\begin{table*}[!t]
    \centering
    \input tables/max.tex
    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios.}\label{tab:max-rewards}
\end{table*}

@@

\subsection{Results}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of normalized rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

\paragraph{Dispersion}
In \Cref{fig:experiments}, the first row displays results for the dispersion scenario (non-differentiable, thus SHAC is not applicable). In the first column (single agent, see also \Cref{tab:max-rewards} and \Cref{apx:fig:experiments-mlp}), both PPO and \fname{} perform well, though \fname{} converges faster thanks to gradient approximation. In the second column (3 agents), PPO fails to converge, whereas \fname{} converges in fewer than 10,000 episodes, suggesting emergent cooperative behavior. In the third column (5 agents), PPO fails to converge, while \fname{} achieves high rewards (with no early stopping) but shows high variance due to increased environment complexity.

\paragraph{Discovery} 
The second row of \Cref{fig:experiments} shows a scenario with partial observability, namely discovery. In the first column (single agent), \fname{} outperforms PPO, though it does not reach the maximum reward, suggesting the world model struggles to capture dynamics. In the second and third columns (3 and 5 agents), PPO again fails to converge, whereas \fname{} maintains superior results, indicating cooperative behavior.

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance. \fname{} converges faster, but SHAC is more consistent due to direct gradient stability. In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes, with SHAC again more stable. 
With more agents, convergence is faster because the task becomes easier (more agents can push the package).
The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

\paragraph{Sampling}
The fourth row of \Cref{fig:experiments} illustrates the sampling scenario, which is differentiable and thus applicable to SHAC. In the first column (single agent), PPO fails to converge, while \fname{} outperforms SHAC. In the second and third columns (3 and 5 agents), PPO again fails to converge, while SHAC and \fname{} both complete training in fewer than 5,000 episodes. Notably, SHAC does not exhibit the same stability advantage it displayed in other scenarios, possibly because gradients derived directly from the environment are noisier compared to those learned by the transition network.
